---
title: "Figure-6-code"
output: html_notebook
---


```{r Load packages, warning=FALSE, message=FALSE}
library(UCell)
library(ggpubr)
library(Seurat)
library(SeuratDisk)
library(tidyverse)
library(assertthat)
library(cowplot)
library(tidyquant)
library(viridis)
library(EnhancedVolcano)
library(MOFA2)
library(HDF5Array)
library(ComplexHeatmap)
library(circlize)
library(clusterProfiler)
library(SPATA2)
library(NICHES)
library(tidyquant)
source(file.path("R","functions.R"))
```

### Figure 6b
```{r  Figure F6b, warning=FALSE,message=FALSE}
load(file.path("data","multimodal_harmony_integrated_gb.RData"))

all <- refquery_hm #LoadH5Seurat(file.path("data","multimodal_harmony_integrated_gb.H5Seurat"))
rm(refquery_hm)

## cell cycle scoring
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Create our Seurat object and complete the initalization steps
all <- CellCycleScoring(all, s.features = s.genes, g2m.features = g2m.genes)
cc_scores <- ScoreSignatures_UCell(all[["RNA"]]@counts, features=list(s.features = s.genes, g2m.features = g2m.genes))
all <- AddMetaData(all, as.data.frame(cc_scores))

## adjust cluster order
order_clusters <- data.frame(seurat_clusters= all$seurat_clusters, row.names = rownames(all[[]])) %>%
  bind_cols(as.data.frame(t(all[["RNA"]]@scale.data))) %>%
  group_by(seurat_clusters) %>%
  summarize_all(.funs=mean) %>%
  as.data.frame()

## adjust cluster order
order_clusters <- data.frame(seurat_clusters= all$seurat_clusters, row.names = rownames(all[[]])) %>%
  bind_cols(as.data.frame(t(all[["RNA"]]@scale.data))) %>%
  group_by(seurat_clusters) %>%
  summarize_all(.funs=mean) %>%
  as.data.frame()
rownames(order_clusters) <- order_clusters$seurat_clusters
order_clusters <- order_clusters$seurat_clusters[hclust(dist(order_clusters[,-1]))$order]

levels(all) <- rev(order_clusters)[c(12,14,13,15:30,11:1)]

## adjust cell types
celltype_cor <- case_when(
  all$seurat_clusters %in% c("19","29") ~ "Neuroectodermal",
  all$seurat_clusters %in% c("17") ~ "CAMs",
  all$seurat_clusters %in% c("3","7","11","14","16","0","12","1","5") ~ "MG",
  all$seurat_clusters %in% c("15","25","4") & all$diagnosis == "GB" ~ "moTAMs",
  all$seurat_clusters %in% c("2") & all$diagnosis == "GB" ~ "mgTAMs",
  all$seurat_clusters %in% c("8") ~ "Class. Mono",
  all$predicted.celltype == "CD14 Mono" ~ "Class. Mono",
  all$predicted.celltype == "CD16 Mono" & all$seurat_clusters == ("23") ~ "Non-Class. Mono",
  T ~ all$predicted.celltype
)
names(celltype_cor) <- colnames(all)
all$celltype_cor <- celltype_cor

## plot clusters
DimPlot(all, label = T, pt.size = 1, repel = T) +
  scale_color_manual(values = c(colors_many, colors,unname(palette_green()))[-c(14,21,28)]) +
  theme_void() +
  NoLegend() +
  labs(title = element_blank())
```
### Figure 6c - top panel
```{r Figure F6c top, warning=FALSE,message=FALSE}
DimPlot(all, group.by = "diagnosis",  label = T) +
  #scale_color_tq(theme = "dark") +
  scale_color_manual(values = unname(palette_green())[c(5,7)]) + 
  theme_void() +
  NoLegend() +
  labs(title = element_blank())
```

### Figure 6c - bottom panel
```{r Figure F6c bottom, warning=FALSE,message=FALSE}
all[[]] %>% 
  mutate(seurat_clusters=factor(seurat_clusters, levels=levels(all))) %>% 
  mosaicGG2(X="seurat_clusters",FILL="diagnosis", colors = unname(palette_green())[c(5,7)])

all[[]] %>% 
  hyper_test_n(var1="seurat_clusters",var2="diagnosis")
```

### Figure 6d - top panel
```{r Figure F6d top, warning=FALSE,message=FALSE}
DimPlot(all, group.by = "compartment", label = T) +
  scale_color_manual(values = unname(palette_dark())[c(2,4)]) + 
  theme_void() +
  NoLegend() +
  labs(title = element_blank())
```


### Figure 6d - bottom panel
```{r Figure F6d bottom, warning=FALSE,message=FALSE}
all[[]] %>% 
  mutate(seurat_clusters=factor(seurat_clusters, levels=levels(all))) %>% 
  mosaicGG2(X="seurat_clusters",FILL="compartment", colors = unname(palette_dark())[c(2,4)]) 

all[[]] %>% 
  hyper_test_n(var1="seurat_clusters",var2="diagnosis")
```

```{r source data F6b_d, echo=FALSE,warning=FALSE,message=FALSE}
### export source data figure 4b
write.csv(data.frame("UMAP_1"=all@reductions$umap@cell.embeddings[,1],
                     "UMAP_2"=all@reductions$umap@cell.embeddings[,2],
                     "Cluster"=all$seurat_clusters,
                     "Diagnosis"=all$diagnosis,
                     "Compartment"=all$compartment ,
                     "Celltype"=all$celltype_cor),
          file.path("data","source_data","panel_F6b_d.csv"))
```

### Figure 6e
```{r Figure F6e, warning=FALSE,message=FALSE}
## subset for tissue resident myeloid cells
gb_mf <- subset(all, subset=celltype_cor %in% c("MG","CAMs","TAMs","Trans. moTAMs") & diagnosis == "GB") #,"MG"

comp_celltype <- case_when(grepl("MG",gb_mf$celltype_cor) ~ "MG",
                          T ~ paste(gb_mf$celltype_cor,gb_mf$compartment,sep="_"))

names(comp_celltype) <- colnames(gb_mf)
gb_mf$comp_celltype <- comp_celltype
Idents(gb_mf) <- gb_mf$comp_celltype

## create the mofa object
if (!file.exists(file.path("data","mofa_model_gb_myeloid_cells.h5"))) {
  ## create MOFA object
  #DefaultAssay(gb_mf) <- "RNA"
  gb_mf <- gb_mf %>% NormalizeData() %>% FindVariableFeatures(nfeatures = 5000) %>% ScaleData() 
  
  MOFAobject_all <- create_mofa(gb_mf, groups = "comp_celltype", features = VariableFeatures(gb_mf), assay="RNA", extract_metadata = T)
  
  ## plot data overview
  plot_data_overview(MOFAobject_all)
  
  ## Default data options
  data_opts <- get_default_data_options(MOFAobject_all)
  
  ## Default model options
  model_opts <- get_default_model_options(MOFAobject_all)
  
  ## Training options
  train_opts <- get_default_training_options(MOFAobject_all)
  train_opts$convergence_mode <- "fast"
  train_opts$seed <- 42
  model_opts$num_factors <- 10
  
  #prepare mofa object
  MOFAobject_all <- prepare_mofa(
    object = MOFAobject_all,
    data_options = data_opts,
    model_options = model_opts,
    training_options = train_opts
  )
  
  ## generate a hdf5 file for the output
  rhdf5::h5createFile(file.path("data","mofa_model_gb_myeloid_cells.h5"))
  
  ## train model
  MOFAobject_all <- run_mofa(MOFAobject_all, outfile = file.path("data","mofa_model_gb_myeloid_cells.h5"), use_basilisk = T)
} else {
  MOFAobject_all <- load_model(file.path("data","mofa_model_gb_myeloid_cells.h5"))
}

## plot model

plt <- plot_variance_explained(MOFAobject_all, x="view", y="factor") +
  geom_text(aes(label=round(value,2))) +
  coord_fixed()

plt

```

```{r source data F6e, echo=FALSE,warning=FALSE,message=FALSE}
### export source data figure 4b
write.csv(data.frame(plt$data),
          file.path("data","source_data","panel_F6e.csv"))
```

```{r Figure F6f, warning=FALSE,message=FALSE}
## gsea cluster profiler
if (!file.exists(file.path("data","mofa_weights2_>.1_gb_mf.csv"))) {
  ## gsea cluster profiler
  mofa_weights2 <- get_weights(MOFAobject_all, views = "RNA", factors = 1:8, 
                               as.data.frame = TRUE) %>% 
    group_by(factor) %>% 
    top_n(100, wt=value)
  #filter(value>.1)
  
  ## convert gene names
  genes <- bitr(unique(mofa_weights2$feature), fromType = "SYMBOL", toType = "ENTREZID", OrgDb = 'org.Hs.eg.db')
  colnames(genes)[1] <- "feature"
  
  ## comparison between clusters
  mofa_weights2 <- mofa_weights2 %>% 
    left_join(genes[!duplicated(genes$feature),]) %>% 
    na.omit()
  
  ## export mofa_weight2
  write.csv(mofa_weights2, file.path("data","mofa_weights2_>.1_gb_mf.csv"))
} else {
  mofa_weights2 <- read.csv(file.path("data","mofa_weights2_>.1_gb_mf.csv"), row.names = 1)
} 

### note that with the current version of clusterProfiler the plot will look slightly different
diff_go_bp <- compareCluster(ENTREZID ~ factor,
                             data=mofa_weights2, 
                             fun = enrichGO,
                             OrgDb = 'org.Hs.eg.db',
                             ont = "BP")

plt <- clusterProfiler::dotplot(diff_go_bp)

plt
```

```{r source data F6f, echo=FALSE,warning=FALSE,message=FALSE}
### export source data figure 4b
write.csv(mofa_weights2,
          file.path("data","source_data","panel_F6f.csv"))
```

### Preprocess the data for Figure 6g - get the reference data
```{r Figure 6g preprocess, warning=FALSE,message=FALSE}

if (!file.exists(file.path("data","GBM1_left.RData"))) {## object 8 contains leptomeninges and cortex
  ## subset it and create spatial segments
  coord_df <- read.csv(file.path("data","Out_GBM1_Scale50_High_Prior_s50_sd50_conf08","b1hi_ctrl_coord_prior08.csv"))
  colnames(coord_df) <- c("barcodes","x","y")
  count_mtr  <- read.csv(file.path("data","Out_GBM1_Scale50_High_Prior_s50_sd50_conf08","segmentation_counts_filt_prior08.csv"),row.names = 1)
  count_mtr <- count_mtr[, colSums(count_mtr) >= 5]
  
  ## subset coord_df for the left half
  coords_GBM1_left <- coord_df[coord_df$x < 12095,]
  count_mtr_GBM1_left <- Matrix(as.matrix(count_mtr[,colnames(count_mtr) %in% coords_GBM1_left$barcodes]), sparse = T)       
  barcodes_GBM1_left <- colnames(count_mtr_GBM1_left)[colnames(count_mtr_GBM1_left) %in% coords_GBM1_left$barcodes]
  coords_GBM1_left <- coords_GBM1_left[coords_GBM1_left$barcodes %in% barcodes_GBM1_left,]
  
  GBM1_left <- initiateSpataObject_CountMtr(
    coords_df = coords_GBM1_left, # your coordinate data.frame
    count_mtr = count_mtr_GBM1_left, # a matrix with unprocessed count values
    sample_name = "CtrlCtx4_GBM1_left",
    image = file.path("data","Out_GBM1_Scale50_High_Prior_s50_sd50_conf08","")
  )
  GBM1_left <- createTrajectories(object = GBM1_left )
  GBM1_left <- createSegmentation(object = GBM1_left )
  
  save(GBM1_left, file = file.path("data","GBM1_left.RData"))
  
  ## subset coord_df for the right half
  coords_GBM1_right <- coord_df[coord_df$x > 12095,]
  count_mtr_GBM1_right <- Matrix(as.matrix(count_mtr[,colnames(count_mtr) %in% coords_GBM1_right$barcodes]), sparse = T)       
  barcodes_GBM1_right <- colnames(count_mtr_GBM1_right)[colnames(count_mtr_GBM1_right) %in% coords_GBM1_right$barcodes]
  coords_GBM1_right <- coords_GBM1_right[coords_GBM1_right$barcodes %in% barcodes_GBM1_right,]
  
  GBM1_right <- initiateSpataObject_CountMtr(
    coords_df = coords_GBM1_right, # your coordinate data.frame
    count_mtr = count_mtr_GBM1_right, # a matrix with unprocessed count values
    sample_name = "CtrlCtx4_GBM1_right",
    image = "/Users/romansankowski/Documents/single_cell_analysis/Cartana_spatial_2022/data/DAPI_50%_scaled/DAPI_Ctrl Ctx 4_ISAP13586301_20220629.png"
  )
  #GBM1_right <- createTrajectories(object = GBM1_right )
  GBM1_right <- createSegmentation(object = GBM1_right )
  GBM1_right <- createTrajectories(object = GBM1_right )
  save(GBM1_right, file = file.path("data","GBM1_right.RData"))
  
} else {
  load(file.path("data","GBM1_left.RData"))
  load(file.path("data","GBM1_right.RData"))
  
}

## setup seurat object
if (!file.exists(file.path("data","seurat_gbm1_left.H5Seurat"))) {
  GBM1_left_seurat <- GBM1_left %>% 
    transformSpataToSeurat
  
  GBM1_right_seurat <- GBM1_right %>% 
    transformSpataToSeurat
  
  SaveH5Seurat(GBM1_left_seurat, file.path("data","seurat_gbm1_left.H5Seurat"))
  SaveH5Seurat(GBM1_right_seurat, file.path("data","seurat_gbm1_right.H5Seurat"))
} else {
  GBM1_left_seurat <-  LoadH5Seurat(file.path("data","seurat_gbm1_left.H5Seurat"))
  GBM1_right_seurat <- LoadH5Seurat(file.path("data","seurat_gbm1_right.H5Seurat"))
}

## assign cell types
## plot celltypes

## merge both objects
all <- merge(x=GBM1_left_seurat, GBM1_right_seurat)

VariableFeatures(all) <- rownames(all) 
all <- all %>% 
  NormalizeData() %>% 
  ScaleData() %>% 
  RunPCA()

## load reference
reference <- LoadH5Seurat(file.path("data","reference_seurat_gb.H5Seurat"))
  
## downsample reference
set.seed(79106)
cell_ids <- reference[[]] %>% 
  rownames_to_column(var = "cellID") %>% 
  group_by(celltype_cor) %>% 
  slice_sample(n=200, replace = T) %>% 
  distinct(cellID, celltype_cor)

## subset for genes present in the nanostring dataset
reference <- reference[rownames(all),cell_ids$cellID]

## remove cells with few counts
reference <- reference[,colSums(reference[["RNA"]]@counts)>10]

## Normalize data
VariableFeatures(reference) <- rownames(reference)
DefaultAssay(reference) <- "RNA"
reference <- reference %>% 
  NormalizeData() %>% 
  ScaleData() %>% 
  RunPCA()

## classify mvp cells
if (!file.exists(file.path("data","GB1_celltype_predictions.RData"))) {
  ## map celltypes
  anchors_GB1 <- FindTransferAnchors(reference = reference , query = all,
                                     dims = 1:30, reference.reduction = "pca")
  predictions_GB1 <- TransferData(anchorset = anchors_GB1, refdata = reference$celltype_cor,
                                  dims = 1:30, k.weight = 20)
  
  ## adjust cell labels
  predictions_GB1$predicted.id_cor <- case_when(predictions_GB1$predicted.id == "Transitory Mo-TAM" ~ "Tr. moTAMs",
                                                     grepl("Mo-TAM", predictions_GB1$predicted.id) ~ "moTAMs",
                                                     grepl("TAM", predictions_GB1$predicted.id) ~ "mgTAMs",
                                                     T ~ predictions_GB1$predicted.id)
  
  save(predictions_GB1, file = file.path("data","GB1_celltype_predictions.RData"))
} else {
  load(file.path("data","GB1_celltype_predictions.RData"))
}



```

### Plot figure 6g - top panel
```{r Figure 6g top panel, warning=FALSE,message=FALSE}
## add metadata 
all <- AddMetaData(all,predictions_GB1)

## ## plot
coord <- rbind(GBM1_left@coordinates[[1]],
               GBM1_right@coordinates[[1]])
rownames(coord) <- coord$barcodes

## reorder based on the seurat object
coord <- coord[colnames(all),]

#rownames(coord) <- gsub("cell","X",coord$barcodes)

coord$segmentation <- gsub("[1-9]$","",all[[]][rownames(coord),]$segmentation)
coord$segmentation_detail <- all[[]][rownames(coord),]$segmentation
coord$celltype_cor <- all[[]][rownames(coord),]$predicted.id_cor

## define colors
cell_cols <- unname(palette_light())
names(cell_cols) <- c("Astro","CAMs","Endothelial cells","Fibroblast","Lymphocytes","mgTAMs","Mural cells","Neuron","Oligo","Tumor","moTAMs","Tr. moTAMs")

coord %>% 
  ggplot(aes(x,y,color=celltype_cor)) +
  geom_point(pch=20, size=.001) +
  scale_color_manual(values = cell_cols[unique(coord$celltype_cor)]) +
  theme_void() +
  coord_fixed() 

```

### Plot Figure 6g - bottom panel
```{r Figure 6g left panel, warning=FALSE,message=FALSE}
coord <- coord %>% 
  mutate(segmentation_detail= ifelse(segmentation_detail=="none", "Solid Tumor", segmentation_detail)) 
coord %>% 
  ggplot(aes(x,y,color=segmentation_detail)) +
  geom_point(pch=20, size=.001) +
  #scale_color_manual(values = cell_cols[unique(coord$celltype_cor)]) +
  scale_color_brewer(palette="Set3") +
  theme_void() +
  coord_fixed() 

```


### Figure 6h - left panel
```{r Figure 6h left panel, warning=FALSE,message=FALSE}
coord %>% 
  filter(segmentation_detail=="hypoxia9") %>% 
  ggplot(aes(x,y,color=celltype_cor)) +
  geom_point(pch=20, size=2) +
  scale_color_manual(values = cell_cols[unique(coord$celltype_cor)]) +
  theme_void() +
  coord_fixed() +
  NoLegend() +
  geom_segment(aes(x = max(x)-550, y = min(y)+50, xend = max(x)-50, yend = min(y)+50), size=1, color="black") +
  geom_text(aes(max(x)-310, y = min(y)+130,label="500 px"))

```

```{r source data F6g_i, echo=FALSE, message=FALSE,warning=FALSE}
write.csv(coord, file.path("data","source_data","panel_F6g_i.csv"))
```


### Figure 6h - middle panel
```{r Figure 6h middle panel, warning=FALSE,message=FALSE}

## assess genes along the trajectory
all_genes <- getGenes(GBM1_right)
atdf_genes <- assessTrajectoryTrends(object = GBM1_right, 
                                     trajectory_name = "hypoxia_left_short", 
                                     variables = all_genes)

# output example
atdf_genes

## ascending genes
ascending_genes <-
  filterTrajectoryTrends(atdf = atdf_genes,
                         limit = 68,
                         trends = "Gradient ascending", 
                         variables_only = FALSE) %>% # return a data.frame
  top_n(-10, wt=auc)

ascending_genes

descending_genes <-
  filterTrajectoryTrends(atdf = atdf_genes,
                         limit = 68,
                         trends = c("Gradient ascending", "Gradient descending"), 
                         variables_only = FALSE) %>% # return a data.frame
  top_n(-10, wt=auc)

descending_genes
descending_genes_vec <- descending_genes$variables

## plot heatmap
hm_colors <- viridis::inferno(n = 100)

plotTrajectoryHeatmap(object = GBM1_right, 
                      trajectory_name = "hypoxia_left_short",
                      variables = descending_genes_vec,
                      arrange_rows = "maxima",
                      colors = hm_colors,
                      show_rownames = TRUE,
                      split_columns = FALSE, 
                      smooth_span = 0.5)

```

### Figure 6i
```{r Figure 6i, warning=FALSE,message=FALSE}
## plot composition of hypoxia 9
coord3 <- coord %>% 
  group_by(segmentation_detail, celltype_cor) %>% 
  summarise(cell_count=n()) %>% 
  group_by(segmentation_detail, .drop = F) %>% 
  mutate(total_cells = sum(cell_count),
         cell_pct=round(cell_count/total_cells*100,2),
         label=paste0(celltype_cor," (",cell_pct,"%)"),
         Cell_ordered=reorder(celltype_cor, cell_pct),
         seg_total=paste0(segmentation_detail, "\n(n=",total_cells,")")) 


coord3 %>% 
  filter(segmentation_detail %in% c("none", "hypoxia9")) %>% 
  ggplot(aes(x=seg_total, y=cell_pct, fill=Cell_ordered, label = label)) +
  geom_col(position = "stack", width = .5,alpha=.8, color="black",lwd =.1) +
  scale_y_continuous(breaks = c(0,50,100)) +
  #scale_fill_tq(guide_none()) +
  scale_fill_manual(values = cell_cols[unique(coord$celltype_cor)]) +
  geom_text_repel(position = position_stack_and_nudge(vjust=.5,y=3), max.overlaps = 30) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size=15)) +
  coord_flip() +
  labs(y="Cell percentage [%]",x=element_blank())

```
### preprocess data for Figure 6j
```{r Figure 6j preprocess, warning=FALSE,message=FALSE}
load(file.path("data","gb_necrosis.RData"))

DefaultAssay(gb_necrosis) <- "Nanostring"
VariableFeatures(gb_necrosis) <- rownames(gb_necrosis)

gb_necrosis <- gb_necrosis %>% 
  NormalizeData() %>% 
  ScaleData() %>% 
  RunPCA()


## classify necrosis cells
if (!file.exists(file.path("data","GBM_1_18_celltype_predictions.RData"))) {
   ## load annotated references single-nuc-seq dataset
  reference <- LoadH5Seurat(file.path("data","reference_seurat.H5Seurat"))

  ## downsample reference
  set.seed(79106)
  cell_ids <- reference[[]] %>% 
    rownames_to_column(var = "cellID") %>% 
    group_by(celltype_cor) %>% 
    slice_sample(n=200, replace = T) %>% 
    distinct(cellID, celltype_cor)
  
  ## subset for genes present in the nanostring dataset
  reference <- reference[rownames(gb_mvp),cell_ids$cellID]
  
  ## remove cells with few counts
  reference <- reference[,colSums(reference[["RNA"]]@counts)>10]
  
  ## re-normalize
  VariableFeatures(reference) <- rownames(reference)
  DefaultAssay(reference) <- "RNA"
  reference <- reference %>% 
    NormalizeData() %>% 
    ScaleData() %>% 
    RunPCA()

  ## map celltypes
  anchors_necrosis <- FindTransferAnchors(reference = reference , query = gb_necrosis,
                                          dims = 1:30, reference.reduction = "pca")
  predictions_necrosis <- TransferData(anchorset = anchors_necrosis, refdata = reference$celltype_cor,
                                       dims = 1:30, k.weight = 20)
  
  ## adjust cell labels
  predictions_necrosis$predicted.id_cor <- case_when(predictions_necrosis$predicted.id == "Transitory Mo-TAM" ~ "Tr. moTAMs",
                                                     grepl("Mo-TAM", predictions_necrosis$predicted.id) ~ "moTAMs",
                                                     grepl("TAM", predictions_necrosis$predicted.id) ~ "mgTAMs",
                                                     T ~ predictions_necrosis$predicted.id)
  
  save(predictions_necrosis, file = file.path("data","GBM_1_18_celltype_predictions.RData"))
} else {
  load(file.path("data","GBM_1_18_celltype_predictions.RData"))
}

## add metadata 
gb_necrosis <- AddMetaData(gb_necrosis,predictions_necrosis)



```

### Figure 6j left panel
```{r Figure 6j right, warning=FALSE,message=FALSE}
## plot
cell_cols <- unname(palette_light())
names(cell_cols) <- c("Astro","CAMs","Endothelial cells","Fibroblast","Lymphocytes","mgTAMs","Mural cells","Neuron","Oligo","Tumor","moTAMs","Tr. moTAMs")
ImageDimPlot(gb_necrosis, boundaries = "segmentation",group.by = "predicted.id_cor", coord.fixed = T, size = 10) +
  scale_fill_manual(values = cell_cols[unique(gb_necrosis$predicted.id_cor)])

## celltype bar plot
gb_necrosis_meta <- gb_necrosis[[]] %>% 
  group_by(predicted.id_cor) %>% 
  summarise(cell_count=n()) %>% 
  mutate(total_cells = sum(cell_count),
         cell_pct=round(cell_count/total_cells*100,2),
         label=paste0(predicted.id_cor," (",cell_pct,"%)"),
         Cell_ordered=reorder(predicted.id_cor, cell_pct),
         seg_total=paste0("\n(n=",total_cells,")")) 

```

### Figure 6j right panel
```{r Figure 6j left, warning=FALSE,message=FALSE}
## plot segments
gb_necrosis <- gb_necrosis %>% 
  AddMetaData(read.csv(file.path("data","gb_necrosis_metadata_segmentation.csv"), row.names="barcodes"))

## manually adjust segmentation to follow the gene expression patterns
segmentation <- case_when(gb_necrosis$segmentation == "none" & GetTissueCoordinates(gb_necrosis)$y < -44000 & GetTissueCoordinates(gb_necrosis)$x > -322000 ~ "tumor",
                          gb_necrosis$segmentation == "none" & GetTissueCoordinates(gb_necrosis)$y < -47000 ~ "tumor",
                          gb_necrosis$segmentation == "none" & GetTissueCoordinates(gb_necrosis)$y > -44000 & GetTissueCoordinates(gb_necrosis)$x < -322000 ~ "necrosis",
                          gb_necrosis$segmentation == "none"  ~ "tumor",
                          T ~ gb_necrosis$segmentation )
names(segmentation) <- colnames(gb_necrosis)

gb_necrosis$segmentation_cor <- segmentation

## celltype bar plot per segments
gb_necrosis_meta <- gb_necrosis[[]] %>% 
  mutate(segmentation_cor=factor(segmentation_cor,levels=c("tumor","necrosis"))) %>% 
  group_by(segmentation_cor,predicted.id_cor) %>% 
  summarise(cell_count=n()) %>% 
  group_by(segmentation_cor) %>% 
  mutate(total_cells = sum(cell_count),
         cell_pct=round(cell_count/total_cells*100,2),
         label=paste0("(",cell_pct,"%)"),
         seg_total=paste0("\n(n=",total_cells,")"),
         segmentation_cor2=paste0(segmentation_cor,"\n(n=",total_cells,")")) 


gb_necrosis_meta %>% 
  ggplot(aes(x=segmentation_cor, y=cell_pct, fill=reorder(predicted.id_cor, cell_pct), label = label)) +
  geom_col(position = "stack", width = .5,alpha=.8, color="black",lwd =.1) +
  scale_y_continuous(breaks = c(0,50,100)) +
  #scale_fill_tq(guide_none()) +
  scale_fill_manual(values = cell_cols[unique(gb_necrosis_meta$predicted.id_cor)]) +
  geom_text_repel(position = position_stack_and_nudge(vjust=.5,y=3), max.overlaps = 30) +
  theme_minimal() +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(size=15)) +
  #coord_flip() +
  labs(y="Cell percentage [%]",x=element_blank())
```

```{r source data F6j, echo=FALSE,warning=FALSE,message=FALSE}
coords_gb <- GetTissueCoordinates(gb_necrosis) 
coords_gb$celltype <- gb_necrosis$predicted.id_cor
coords_gb$segmentation <- gb_necrosis$segmentation_cor

write.csv(coords_gb, file.path("data","source_data","panel_F6j.csv"))
```

### Figure 6k
```{r Figure 6k, warning=FALSE,message=FALSE}
load(file.path("data","multimodal_harmony_integrated_gb.RData"))

all <- refquery_hm #LoadH5Seurat(file.path("data","multimodal_harmony_integrated_gb.H5Seurat"))
rm(refquery_hm)


## cell cycle scoring
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes

# Create our Seurat object and complete the initalization steps
all <- CellCycleScoring(all, s.features = s.genes, g2m.features = g2m.genes)
cc_scores <- ScoreSignatures_UCell(all[["RNA"]]@counts, features=list(s.features = s.genes, g2m.features = g2m.genes))
all <- AddMetaData(all, as.data.frame(cc_scores))

## adjust cluster order
order_clusters <- data.frame(seurat_clusters= all$seurat_clusters, row.names = rownames(all[[]])) %>%
  bind_cols(as.data.frame(t(all[["RNA"]]@scale.data))) %>%
  group_by(seurat_clusters) %>%
  summarize_all(.funs=mean) %>%
  as.data.frame()

## adjust cluster order
order_clusters <- data.frame(seurat_clusters= all$seurat_clusters, row.names = rownames(all[[]])) %>%
  bind_cols(as.data.frame(t(all[["RNA"]]@scale.data))) %>%
  group_by(seurat_clusters) %>%
  summarize_all(.funs=mean) %>%
  as.data.frame()
rownames(order_clusters) <- order_clusters$seurat_clusters
order_clusters <- order_clusters$seurat_clusters[hclust(dist(order_clusters[,-1]))$order]

levels(all) <- rev(order_clusters)[c(12,14,13,15:30,11:1)]

## adjust cell types
celltype_cor <- case_when(
  all$seurat_clusters %in% c("19","29") ~ "Neuroectodermal",
  all$seurat_clusters %in% c("17") ~ "CAMs",
  all$seurat_clusters %in% c("3","7","11","14","16","0","12","1","5") ~ "MG",
  all$seurat_clusters %in% c("15") & all$diagnosis == "GB" ~ "Trans. moTAMs",
  all$seurat_clusters %in% c("2","4") & all$diagnosis == "GB" ~ "TAMs",
  #all$seurat_clusters %in% c("2") & all$diagnosis == "GB" ~ "mgTAMs",
  all$seurat_clusters %in% c("8") ~ "Class. Mono",
  all$predicted.celltype == "CD14 Mono" ~ "Class. Mono",
  all$predicted.celltype == "CD16 Mono" & all$seurat_clusters == ("23") ~ "Non-Class. Mono",
  T ~ all$predicted.celltype
)

names(celltype_cor) <- colnames(all)
all$celltype_cor <- celltype_cor

## run adt analysis
all_adt <- all
DefaultAssay(all_adt) <- "ADT"

VariableFeatures(all_adt) <- rownames(all_adt)
all_adt <- all_adt %>% 
  subset(subset = nCount_ADT > 50) %>%  #  %>% 
  NormalizeData(normalization.method = 'CLR', margin = 2) %>% 
  ScaleData(vars.to.regress = "nCount_ADT") %>% 
  RunPCA(reduction.name = 'apca')

## set composite celltypes
composite_celltype <- paste(all_adt$compartment, all_adt$diagnosis, all_adt$celltype_cor,sep="_")

composite_celltype[grepl("(MG|mg)",composite_celltype)] <- gsub("(LM_|CTX_)","",composite_celltype[grepl("(MG|mg)",composite_celltype)])

names(composite_celltype) <- colnames(all_adt)
all_adt$composite_celltype <- composite_celltype

all_adt_myeloid <- all_adt[,grepl("(MG|TAM|CAMs)", all_adt$composite_celltype)]

Idents(all_adt_myeloid) <- all_adt_myeloid$composite_celltype

## both compartments
## differential genes per compartment
myeloid_adt_markers <- FindAllMarkers(all_adt_myeloid,logfc.threshold=.2) 
top5 <- myeloid_adt_markers %>% 
  filter(p_val_adj<.05 & avg_log2FC>0) %>% 
  group_by(cluster) %>% 
  top_n(5, wt=avg_log2FC) 

mean_expr <- data.frame(t(as.matrix(all_adt_myeloid[["ADT"]]@data[unique(top5$gene),])), "cluster"=all_adt_myeloid$composite_celltype) %>%  
  filter(cluster != "CTX_Ctrl_CAMs") %>% ### exclude ctrl cams because they were so few
  group_by(cluster) %>% 
  summarise_all(.funs = mean) %>% 
  as.data.frame()

rownames(mean_expr) <- mean_expr$cluster
mean_expr$cluster <- NULL
prots <- gsub("(Hu\\.|\\.M5E2|HuMs\\.|_RPA_T4)","",colnames(mean_expr))
prots <- gsub("(\\.)","_",prots)
colnames(mean_expr) <- prots
mat <- scale(as.matrix(mean_expr))

## add row annotation 
row_anns <- mat %>%
  as.data.frame() %>% 
  rownames_to_column(var = "sample") %>% 
  mutate(compartment = ifelse(grepl("MG",sample),"CTX",gsub("_.*","",sample)),
         diagnosis = ifelse(grepl("GB",sample), "GB","Ctrl")) %>% 
  column_to_rownames("sample")

row_ha <- rowAnnotation(compartment = row_anns$compartment,
                        diagnosis = row_anns$diagnosis,
                        col = list(compartment = c("LM" = unname(palette_dark())[4], 
                                                 "CTX" = unname(palette_dark())[2]),
                                    diagnosis =  c("Ctrl" = unname(palette_green())[c(5)],
                                                   "GB" = unname(palette_green())[c(7)])),
                        gap = unit(2.5, "mm"),
                        border = TRUE)

Heatmap(mat, cluster_rows = T,
        cluster_columns = T, 
        #right_annotation = row_ha,
        clustering_method_rows = "ward.D2",
        width = ncol(mat)*unit(5, "mm"), 
        height = nrow(mat)*unit(5, "mm"))
```

```{r source data F6k, echo=FALSE,warning=FALSE,message=FALSE}
write.csv(mat, file.path("data","source_data","panel_F6k.csv"))
```


Roman Sankowski
03-10-2023
email: roman.sankowski@uniklinik-freiburg.de
